---
title: "Lab 03"
author: "STA 198 Summer 2020 (Jiang)"
output:
  rmarkdown::html_document:
    css: "./sta198-labs.css"
    toc: TRUE
    toc_float: TRUE
---
<br>

Today, we will be learning how to use R to calculate probabilities corresponding to events for some of the distributions we covered in class. There is no associated dataset, but I will be referencing some examples you've seen before.

Click on the following link to clone the lab directory into the class workspace: 

[https://classroom.github.com/a/1tOb2_jE](https://classroom.github.com/a/1tOb2_jE)

This assignment is due June 09 at 11:59p via Gradescope (you know the drill). 

As a brief reminder: if you're unsure how a function works or what its arguments are, simply type `?` in front of it and hit enter (`?pbinom` for instance). The "Help" tab will open and provide a summary of the function as well as some examples.

# The binomial distribution

`R` has a variety of functions that make working with distributions much easier. For the binomial distribution, we can... 

- Use the `dbinom` function to calculate the mass at a given point (i.e., the probability of obtaining exactly *k* successes in n independent Bernoulli trials with success probability *p*)
- Use the `pbinom` function to calculate left-tail probabilities (i.e., the probability of obtaining *k* *or fewer* successes)
- Use the `rbinom` function to create random samples (we won't really be using this, but it's included here for completeness)

**Example**: For a random sample of 12 births in the US, what is the probability of observing exactly 3 preterm births, if the births are independent and the probability of a preterm birth is 0.1?

**Answer**: This is a binomial distribution with 12 total draws and success probability of 0.1. We are interested in evaluating P(X = 3), where X follows a binomial distribution with n = 12 and p = 0.1. Note that the syntax is `dbinom(k, n, p)`, for obtaining k successes in n independent Bernoulli trials with success probability p.

```{r binom1}
dbinom(3, 12, 0.1)
```

**Example**: For a random sample of 12 births in the US, what is the probability of observing 3 *or fewer* preterm births, if the births are independent and the probability of a preterm birth is 0.1?

**Answer**: This is a binomial distribution with 12 total draws and success probability of 0.1. We are interested in evaluating P(X ≤ 3), where X follows a binomial distribution with n = 12 and p = 0.1. The syntax is `pbinom(k, n, p)`, for obtaining *k* *or fewer* successes in n independent Bernoulli trials with success probability p.

```{r binom2}
pbinom(3, 12, 0.1)
```

**Example**: For a random sample of 12 births in the US, what is the probability of observing more than 6 preterm births, if the births are independent and the probability of a preterm birth is 0.1?

**Answer**: This is a binomial distribution with 12 total draws and success probability of 0.1. We are interested in evaluating P(X > 6) = 1 - P(X ≤ 6), where X follows a binomial distribution with n = 12 and p = 0.1. 

```{r binom3}
1 - pbinom(6, 12, 0.1)
```

**Example**: Simulate 50 random samples from a binomial distribution where there were 100 total independent Bernoulli draws with success probability 0.4. Save these draws as a vector named `binom_draws`.

**Answer**: 

```{r binom4}
binom_draws <- rbinom(50, 100, 0.4)
```

Note the syntax here. When using `rbinom`, we start with the number of simulated draws we want, then the `n` parameter from the binomial distribution, and then the success probability. Do not get the number of simulated draws confused with the number of Bernoulli draws!

# The Poisson distribution

Similar functions exist for the Poisson distribution. We can...

- Use the `dpois` function to calculate the mass at a given point
- Use the `ppois` function to calculate left-tail probabilities (i.e., the probability of obtaining *k* *or fewer* counts, with a given $\lambda$)
- Use the `rpois` function to create random samples from the Poisson distribution (we won't really be using this, but it's included here for completeness)

**Example**: Suppose the monthly number of ulcerative colitis flare-ups in an IBD patient follows a Poisson distribution with parameter $\lambda$ = 1. What is the probability that this patient will have 3 flare-ups in a given month?

**Answer**: This is a Poisson distribution with parameter $\lambda$ = 1. We are interested in evaluating P(X = 3), where X follows a Poisson($\lambda$) distribution. Note that the syntax is `dpois(k, lambda)`, where $k$ is the count and $\lambda$ is the parameter.

```{r pois1}
dpois(3, 1)
```

**Example**: Suppose the monthly number of ulcerative colitis flare-ups in an IBD patient follows a Poisson distribution with parameter $\lambda$ = 1. What is the probability that this patient will have more than 2 flare-ups in a given month?

**Answer**: This is a Poisson distribution with parameter $\lambda$ = 1. We are interested in evaluating P(X > 2) = 1 - P(X ≤ 2), where X follows a Poisson($\lambda$) distribution. Note that the syntax is `qpois(k, lambda)`.

```{r pois2}
1 - ppois(2, 1)
```

**Example**: Simulate 200 random samples from a Poisson distribution with parameter 2. Save these draws as a vector named `pois_draws`.

**Answer**: 

```{r pois3}
pois_draws <- rpois(200, 2)
```

Note the syntax here. When using `rbinom`, we start with the number of simulated draws we want, then the `n` parameter from the binomial distribution, and then the success probability. Do not get the number of simulated draws confused with the number of Bernoulli draws!

# The normal (Gaussian) distribution

You guessed it, we can do the same for the normal distribution.

- Use the `dnorm` function to calculate the *density* of the normal distribution at a given point
- Use the `pnorm` function to calculate left-tail probabilities (i.e., the area under the curve, to the *left* of the point you choose). This is the integral from $-\infty$ to the point you specify 
- Use the `rnorm` function to create random samples from the normal distribution (we won't really be using this, but it's included here for completeness)

We will also discuss one additional function that is essentially the inverse of `pnorm`:

- Use the `qnorm` function to determine the value corresponding to a specific quantile of the normal distribution

**Note**: when you are using these functions, you must specify the *standard deviation*, not the variance.

**Example**: Suppose the mean bone density for a 65-year old white woman is normally distributed with mean 809 mg/cm$^3$ and standard deviation of 140 mg/cm$^3$. Your bone density is 698 mg/cm$^3$. What percentile are you at for bone density, compared to your peers?

**Answer**: We are interested in evaluating P(X ≤ 698), where X follows a N(809, 140$^2$) distribution. Note that the syntax is `pnorm(value, mean, sd)`.

```{r norm1}
pnorm(698, 809, 140)
```

**Example**: Suppose the mean bone density for a 65-year old white woman is normally distributed with mean 809 mg/cm$^3$ and standard deviation of 140 mg/cm$^3$. Your bone density is 950 mg/cm$^3$. What percent of women have bones that are even denser than yours?

**Answer**: We are interested in evaluating P(X > 950), where X follows a N(809, 140$^2$) distribution. This is equal to 1 - P(X ≤ 950).

```{r norm2}
1 - pnorm(950, 809, 140)
```

**Example**: Suppose the mean bone density for a 65-year old white woman is normally distributed with mean 809 mg/cm$^3$ and standard deviation of 140 mg/cm$^3$. Suppose the bottom 1\% of women in terms of bone density are at extreme risk for fractures. What bone density corresponds to this threshold?

**Answer**: We are interested in calculating the 1st percentile of a N(809, 140$^2$) distirbution. This is given by

```{r norm3}
qnorm(0.01, 809, 140)
```

# Your turn!

1. **5 points** Make sure your GitHub repository for this assignment has at least three commits, each with a meaningful commit message.

**For all exercises, be careful with strict vs. non-strict inequalities.**

According to CDC data, approximately 91% of North Carolina adults regularly use their seatbelt in cars.

2. **3 points** What is the expected number of randomly selected North Carolinians out of 20 that regularly use their seatbelt in cars? (round to the nearest whole number of people)
3. **3 points** Suppose you select a sample of 20 randomly selected North Carolinians. What is the probability that you would find exactly your expected number (from Exercise 2) who regularly wear seatbelts?
4. **3 points** Now suppose you select a sample of 100 randomly selected North Carolinians. What is the expected number of regular seatbelt wears out of 100? 
5. **3 points** What is the probability that you would find exactly that number (that you found in Exercise 4) out of 100 who are regular seatbelt wearers?
6. **3 points** What is the probability that you would find fewer than 85 randomly selected North Carolinians out of 100 who are regular seatbelt wearers?
7. **4 points** Suppose you select a random sample of 40 North Carolinians. What is the probability that you find more than 5 individuals who are NOT regular seatbelt wearers?

Suppose that the number of fatalities due to car accidents in a given county follows the Poisson distribution with a given mean. According to NCDOT data, the 10-year average for Durham county is 23. Assume that this rate is constant and that car crash fatalities are independent (likely violated in practice due to multi-car collisions, but for the sake of these problems, assume they are).

8. **3 points** What is the probability that there are exactly 23 fatalities due to car accidents in a given year?
9. **3 points** What is the probability that Durham experiences more than 30 fatalities due to car accidents in a given year?
10. **3 points** What is the probability that Durham experiences between 20 and 25 fatalities (inclusive) due to car accidents in a given year?
11. **4 points** What is the probability that Durham does not experience any fatalities due to car accidents in a given *month*?

Suppose the mean bone density for a 65-year old white woman is normally distributed with mean 809 mg/cm$^3$ and standard deviation of 140 mg/cm$^3$ (these are fairly accurate numbers).

12. **3 points** What bone density corresponds to the 5th, 10th, 50th, and 99th percentiles?
13. **3 points** The density of solid pine is 600 mg/cm$^3$. What is the probability that a randomly-sampled 65-year old white woman has bone density *less* than that of solid pine?
14. **3 points** The density of solid mahogany is 710 mg/cm$^3$. What is the probability that a randomly-sampled 65-year old white woman has bone density *greater* than that of solid mahogany?
15. **4 points** The density of plywood is 540 mg/cm$^3$. What is the probability that, out of 100 randomly-sampled unrelated 65-year old white women, at least 5 of them have bones less dense than plywood?