---
title: "Lab 01"
author: "STA 198 Summer 2020 (Jiang)"
output:
  rmarkdown::html_document:
    css: "./sta198-labs.css"
    toc: TRUE
    toc_float: TRUE
---
<br>

Today's lab will introduce `ggplot2` as a visualization package to make 
professional-quality graphics. 

This assignment **will** be graded for credit.
You must upload the knitted .pdf document corresponding to your repository to
Gradescope by 11:59pm, US Eastern Time, on May 26 in order to receive credit. 
Part of your lab grade will depend on having a well-structured repository (more
instructions in Exercise 1, below). As a reminder, no late work is accepted.

# Workflow

Remember the following steps in the workflow. If you have any questions or need
a refresher on how to perform them, refer back to Lab B0 or ask a lab TA or Yue.

- Clone an existing assignment repository into a personal repository.
- Create a new RStudio Cloud Project based on your cloned repository in the 
class workspace.
- Perform your data analysis in RStudio Cloud in an R Markdown document 
(staging, committing, and pushing changes to your repository as necessary).
- Knit your final product into a .pdf document, and push one final time to your
repository.
- Upload this .pdf document to Gradescope.

Click on the following link to clone the lab directory: 

[https://classroom.github.com/a/tL1rmUMs](https://classroom.github.com/a/tL1rmUMs)

# The Grammar of Graphics

As a reminder from lecture, The Grammar of Graphics was a concept introduced by
Leland Wilkinson in his book (appropriately titled *The Grammar of Graphics*).
This formal grammar was used to define the components of a data visualization, 
and was adapted into the `tidyverse` in R. 

One of the biggest advantages of ``ggplot2`` as implemented in R is that it has
a very consistent, readable syntax for use in coding. Additionally, since it is
based on a formal grammar, the code itself is very *modular*, with each line
corresponding to a specific element of the intended graph. However, note that
it is still up to the user to decide which graph is appropriate for the 
situation. It is entirely possible to create non-sensical graphs (such as the
line plot connecting states' obesity rates in alphabetical order).

In the guided portion of today's lab, we will be reproducing the plots from the
lecture slides. The graded component will have you creating your own data
visualizations.

# Guided activity

Let's go through some of the layers in the Grammar of Graphics, adding them one
at a time until we achieve a final result. Follow along with the following 
steps, making sure to note what changes upon the addition of each new layer.

## Before we start

**Remember that the Console is separate from the knitting environment.** The
code you type/run in the Console (this includes the green play button) does not
affect the knitted document, and vice-versa. This often comes into play when
loading libraries or defining datasets/variables.

## Loading libraries and data

The `ggplot2` package is part of the `tidyverse`, and is automatically loaded
when the `tidyverse` package is loaded. Since I've already installed it on the
back-end, all that you need to do is load the library in order to use the
functions in it:

```{r packages, echo = TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
```

We'll be using the `cdc` dataset to recreate the plot on Slide 17 in Lecture 02 step-by-step. Import the data with the following code:

```{r cdcdata, echo = TRUE, warning=FALSE, message=FALSE}
cdc <- read_csv("data/cdc.csv")
```

Note the use of the `<-` operator. This is the **assignment operator**, and is
used to assign the value of the right side to the left side. That is, we are
creating an object `cdc` by reading-in the dataset `cdc_cleaned.csv` in the 
`data` folder using the `read_csv` function, which is part of the tidyverse 
(there is a base R equivalent, but `read_csv` is preferred...feel free to ask
me why!).

We can take a look at the dataset by using either of the following two code
options. Using `View()` opens a browsing tab in the top left-window, and using
`glimpse()` provides a brief summary in the output area. The `glimpse()` 
function tells us how many observations and variables there are in the dataset,
the data types of the variable (don't worry about this now), and it also 
provides the first few observations for each variable so you can get a feel for
the dataset.

```{r data-stuff, eval = F}
View(cdc)
glimpse(cdc)
```

## Creating our scatterplot

Run the following code step-by-step and see what changes each time. When 
creating `ggplot2` visualizations, a best practice is to separate each new layer
on its **own separate line**.

Start by 
specifying the data (note that nothing should be plotted):

```{r, eval = F}
ggplot(data = cdc)
```

Next, **map** variables in the dataset to various **aesthetics**. Let's map  
variables in our dataset onto the x- and y-axes:

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity))
```

You might notice that nothing gets plotted -- this is because we haven't yet
specified what type of plot to create. Let's go ahead and specify a **geometry**
corresponding to a scatterplot. This is done by adding a new layer (literally
add it with a + sign on a new line):

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity)) +
  geom_point()
```

We can change global options on how the plot is formed inside the geometry. For
instance, compare the previous plot to the following one:

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity)) +
  geom_point(size = 3, color = "skyblue")
```

What changed? Now, instead of making all points skyblue, let's color code them
by mapping the HDI variable onto the color aesthetic:

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity, color = hdi)) +
  geom_point()
```

Note that if we want the colors of the points to depend on the value of another
variable, we must specify that as an aesthetic mapping. If we want ot make a
global visual change, we can specify that in the geometry.

Now, let's add some labels and a title in a new layer. Note that a caption is 
automatically added:

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity, color = hdi)) +
  geom_point() +
  labs(title = "Adequate aerobic activity associated with lower obesity", 
       x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)")
```

This is in fact the plot presented on Slide 17. To create the plot on Slide 18
where we have **facetted** the points by HDI instead of color coding, we can
add in a faceting layer, where we facet by HDI instead of mapping it to the
color aesthetic:

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity)) +
  geom_point() +
  facet_grid(.~hdi) + 
  labs(title = "Adequate aerobic activity associated with lower obesity", 
       x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)")
```

We can also change the direction of faceting. For instance, try the following
code and compare it to your previous plot. Which is more effective in telling
a story, and why?

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = exercise, y = obesity)) +
  geom_point() +
  facet_grid(hdi~.) + 
  labs(title = "Adequate aerobic activity associated with lower obesity", 
       x = "Adequate aerobic activity (%)", 
       y = "Obesity (%)")
```

Note that there are a lot of other options and layers we can include, but this
is just a basic introduction. 

## The other plots from class...

This section has the other basic plots introduced in Lecture 02. Take the time
to reproduce them and see what each line of code is doing:

```{r, eval = F}
ggplot(data = cdc, mapping = aes(x = region)) + 
  geom_bar(fill = "steelblue") + 
  labs(title = "The South has the largest number of states", 
       x = "Region", 
       y = "Count")
```

```{r, eval = F}
# Note: if you don't want to separate by HDI, get rid of the x mapping
ggplot(data = cdc, mapping = aes(x = hdi, y = obesity)) + 
  geom_boxplot() + 
  labs(title = "Lower HDIs associated with higher obesity %", 
       x = "HDI", 
       y = "Adult Obesity (%)")
```

```{r, eval = F}
ggplot(data = dat, mapping = aes(x = obesity)) +
  geom_histogram(color = "darkblue", fill = "lightblue", binwidth = 2)+ 
  labs(title = "Most states have approx. 30% adult obesity", 
       x = "Adult Obesity (%)", 
       y = "Count")
```

```{r, eval = F}
# Note: I've taken only the first four rows of the cdc dataset...more details
#       regarding data manipulation will be provided in Lab 02. 
ggplot(data = cdc[1:4,], mapping = aes(x = state, y = obesity, group = 1)) +
  geom_line() + 
  geom_point() + 
  labs(title = "Obesity by State", 
       x = "State", 
       y = "Adult Obesity (%)")
```

# Additional plotting resources

RStudio has written a great ``ggplot2`` cheatsheet linked [here](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf).
It's definitely a lot to take in, but it has virtually everything you'd ever 
need in one page. Selva Prabhakaran wrote an in-depth, three-part tutorial 
available on his website [here](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html),
and I definitely encourage you to take a look in your free time. The third part
of his tutorial lists 50 visualization examples of all types, as well as the 
code needed to create them.

# Your turn

Parkinson's disease is a progressive, incurable neurological disease that leads 
to motor dysfunction and potential cognitive decline, dementia, and reduced life 
expectancy. It is important to diagnose Parkinson's disease early in order to 
start effective disease management.

Researchers at Oxford and the National Center for Voice and Speech in Denver 
made voice measurements on a number of individuals to see whether they could 
diagnose Parkinson's disease based on acoustic signals in peoples' voices using 
spectral analysis. They made a number of measurements among healthy and 
Parkinson's disease patients. 

The dataset in your repository has been adapted from Little et al., Biomed. Eng. 
Online (2007). It is composed of diferent voice measurements, with each 
observation corresponding to a voice recording from a participant.

Of interest are the following variables:

- `avg.f.q` average vocal fundamental frequency divided into groups: 1 (low), 2 (medium), 3 (high)
- `jitter` jitter (in %), a measure of fundamental frequency variation
- `shimmer` shimmer (in dB), a measure of amplitude variation
- `hnr` the ratio of tonal to noise components in the sound clip
- `status` disease status

**Remember to give all graphs effective titles and axis labels.**

1. **6 points** Make sure your GitHub repository for this assignment has at 
least three commits, each with a meaningful commit message.

2. **4 points** Load the `tidyverse` into R, and then read the `parkinsons.csv` 
file using the `read_csv()` function (giving your dataset a meaningful name). 
How many observations and variables are in this dataset?

3. **10 points** Provide a graph that visualizes the average vocal fundamental 
frequency of these voice clips.

4. **15 points** Provide a graph that visualizes the amplitude variation of 
these voice clips, separated by disease status. Describe the distributions; what 
do you notice?

5. **15 points** Provide a graph that visualizes the relationship between 
amplitude variation and fundamental frequency variation in these voice clips, 
and color-code the points by disease status. Provide a legend for the color 
along the bottom of your plot. Describe this relationship; what do you notice?
<br><br>**Hint**: use a `theme()` layer to control where the legend is 
positioned. If you want to look at the help file or documentation for any R
function, you can use `help()` or `?`. For example, if you wanted to look at the 
documentation for the `theme()` function, type in `help(theme)` or `?theme` into 
the Console. 

You must turn in a knitted .pdf file to Gradescope from a markdown file based on 
the template in your cloned repository by 11:59p US Eastern on May 26th in order 
to receive credit. Be sure to "associate" questions appropriately on Gradescope. 
As a reminder, late work is not accepted.
